### 2015-01-05 doc/thesis/evaluation.tex
Minor improvements to the text about typing Lua Standard Library and MD5

There are three case studies that we randomly selected from the LuaRocks
repository, they are: HTTP Digest \citep{luahttpdigest},
Typical \citep{luatypical}, and Mod 11 \citep{luamod11}.
The first provides client side HTTP digest authentication for Lua.
The second is an extension to the primitive function \texttt{type}.
The third is a generator and checker of modulo 11 numbers.
We randomly selected these libraries to evaluate Typed Lua for
rewriting existing Lua modules.

@@

\begin{table}[!ht]
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|c|c|}
\hline
\textbf{Case study} & \textbf{easy} & \textbf{poly} & \textbf{inter} & \textbf{hard} & \textbf{Total} & \textbf{\%} \\
\hline
Lua Standard Library & 80 & 6 & 9 & 34 & 129 & 62\% \\
\hline
MD5 & 13 & 0 & 0 & 0 & 13 & 100\% \\
\hline
LuaSocket & 109 & 1 & 2 & 11 & 123 & 89\% \\
\hline
HTTP Digest & 0 & 0 & 1 & 0 & 1 & 0\% \\
\hline
Typical & 1 & 0 & 0 & 0 & 1 & 100\% \\
\hline
Modulo 11 & 7 & 0 & 0 & 2 & 9 & 78\% \\
\hline
Typed Lua Compiler & 144 & 0 & 1 & 9 & 154 & 94\% \\
\hline
\end{tabular}
\end{center}
\caption{Evaluation results for each case study}
\label{tab:evalbycase}
\end{table}

@@

In the following sections we discuss each case study in more detail.
For each case study, we split the evaluation results according
to the modules that each one of them includes.
This separation by case study allows us to better discuss the
contributions and limitations of our type system.

@@

All of the modules in the standard library are implemented in C, so
we used Typed Lua to type just the interface of each module.
The \texttt{debug} module is the only one that we did not include in our
evaluation results, because it provides several functions that violate
basic assumptions about Lua code \citep{luamanual}.
For instance, we can use its function \texttt{setlocal} to change the value
of a local variable that is not visible in the current scope.
Table \ref{tab:evallsl} summarizes the evaluation results for the Lua Standard Library.

\begin{table}[!ht]
\begin{center}
\begin{tabular}{|c|c|c|c|c|c|c|}
\hline
\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{inter} & \textbf{hard} & \textbf{Total} & \textbf{\%} \\
\hline
base & 9 & 1 & 3 & 13 & 26 & 35\% \\
\hline
coroutine & 1 & 0 & 0 & 6 & 7 & 14\% \\
\hline
package & 5 & 0 & 0 & 3 & 8 & 62\% \\
\hline
string & 12 & 0 & 0 & 4 & 16 & 75\% \\
\hline
utf8 & 6 & 0 & 0 & 0 & 6 & 100\% \\
\hline
table & 1 & 5 & 0 & 1 & 7 & 14\% \\
\hline
math & 22 & 0 & 4 & 1 & 27 & 81\% \\
\hline
io & 15 & 0 & 0 & 6 & 21 & 71\% \\
\hline
os & 9 & 0 & 2 & 0 & 11 & 82\% \\
\hline
\end{tabular}
\end{center}
\caption{Evaluation results for Lua Standard Library}
\label{tab:evallsl}
\end{table}

We could give precise static types to most of the modules,
but some of them are still hard to type with the current type
system.
Now, we will discuss why some modules are difficult to type,
and what kind of limitations they represent on our type system.

The \texttt{base} module was quite hard to type because it
includes several functions that have a dynamic behavior.
For instance, the function \texttt{next} traverses all the
fields of a table, but the order that the indices are
enumerated is not specified.
For this reason, it is also difficult to type the function
\texttt{pairs}.
The functions \texttt{getmetatable}, \texttt{rawget}, and
\texttt{rawset} are also examples of functions that their
behavior depend on the C implementation.

There are also some functions that have a dynamic behavior and
their return type depend on their input type.
For instance, the functions \texttt{assert} and \texttt{select}
can return all the values that they received as their arguments.

We could give a more precise type to \texttt{ipairs} if our
type system had parametric polymorphism, as this function
returns an iterator function that operates over a list of
elements.

Furthermore, the \texttt{base} module also includes two very
special cases: the functions \texttt{tonumber} and \texttt{collectgarbage}.

The function \texttt{tonumber} is a special case because the
type of one input parameter depends on the type of another.
For instance, there are two static types that we can
assign to the function \texttt{tonumber}:
\texttt{(value) -> (number?)} and
\texttt{(string, number) -> (number?)}.
More precisely, the first argument of \texttt{tonumber} can be
a value of any type if it is the only argument, but it must
be a value of type string if there is a second argument,
which must be a value of type number.

The function \texttt{collectgarbage} is a special case because
its return type depends on the value of the first parameter,
which should be a string.
This function can return one number, one boolean, or two numbers
depending on the literal string that was given as its first argument.

We could not type the \texttt{coroutine} module because our
type system cannot describe the computational effects of a program.
Lua has one-shot delimited continuations \citep{james2011yield}
in the form of \emph{coroutines} \citep{moura2009rc}, and
effect systems \citep{nielson1999type} are an approach that we
could use to describe control transfers with continuations.
However, for now, coroutines are out of the scope of our type
system, and we use an empty \emph{userdata} declaration
to represent the type \emph{thread}.

@@

We could type all the members of the \texttt{string} module.
Lua programmers often use the functions of the \texttt{string} module
in an object-oriented style, and we treat this feature as a
special case in the implementation.
For this reason, we did not include the typing of the \texttt{string}
methods in our evaluation results.

@@

However, the lack of parametric polymorphism did not prevent us from
giving a precise type to \texttt{table.concat}, as it operates over lists
where all elements are strings or numbers, and we can use the table type
\texttt{\{string|number\}} to express this type on our type system.

Even if our type system had parametric polymorphism, it would
be difficult to type \texttt{table.insert}, as its type depends on
the calling arity.
We can call \texttt{table.insert} passing two or three parameters.
The first parameter is always a list.
If we call it with two arguments, then the second parameter
is the value to be inserted at the end of the list.
If we call it with three arguments, then the third parameter
is the value to be inserted in the list, and the second
parameter is the position where it should be inserted.
This function also does not follow the semantics of Lua on
discarding extra arguments, and generates a run-time error whenever
we pass more than three arguments, even if the first three arguments
match its signature.

Even though the \texttt{math} module is straightforward to type,
it includes a special case that is hard to type: the function \texttt{math.random}.
This function is difficult to type because its type also depends
on the calling arity.
We can call \texttt{math.random} passing zero, one, or two parameters.
If we pass no parameter, then it returns a random floating point
number between \texttt{0} and \texttt{1}.
If we pass one or two parameters, then all the parameters should be
integers and it returns an integer number.
Another problem with this function is that it also does not follow
the semantics of Lua on discarding extra arguments, and generates
a run-time error whenever we pass more than three arguments.
There is also a problem on its documentation, as it suggests that
the two integers are optional parameters, but, for \texttt{math.random},
optional parameters behave in a different way.
Usually, Lua functions replace optional parameters with a default value
when they are \texttt{nil}, but \texttt{math.random} generates a run-time
error instead.

The \texttt{bit32} module was quite straightforward to type,
as its members are only bitwise operations over numbers.

@@

The function \texttt{io.close} is difficult to type
because its return type depends on whether the file handle that is
being closed was created with \texttt{popen} or not.
If it was created with \texttt{popen}, then \texttt{io.close}
has type \texttt{(file?) -> (boolean?, string, number)};
otwerwise \texttt{io.close} has type \texttt{(file?) -> (boolean)?}.
The method \texttt{file:close} is difficult to type for the same reason.

The functions \texttt{io.read} and \texttt{io.lines} are difficult
to type because their return types depend on their input types.
The type of \texttt{io.read} can be either \texttt{() -> (string*)}
or \texttt{(string|number*) -> (string|number*)}, and
the type of \texttt{io.lines} can be either \texttt{(string?) -> (string*)}
or \texttt{(string, string|number*) -> (string|number*)}.
Returning \texttt{(string|number*)} is also painful to programmers, as
they need to constrain the value of the return type among three
different types before cocatenating it, for example.
The methods \texttt{file:read} and \texttt{file:lines} are difficult
to type for the same reason.

The function \texttt{os.execute} is the only member of the \texttt{os}
module that we could not give a precise type.
This function is difficult to type because its return type depend on
the input type.
Its type can be either \texttt{() -> (boolean)} or
\texttt{(string) -> (boolean?, string, number)}.

The evaluation results show that our type system should include
parametric polymorphism, intersection types, and effect types,
as these features would help us increase the static typing of the
Lua Standard Library.
Parametric polymorphism would allow us to define generic function and table types.
Intersection types would allow us to define overloaded function types.
Effect types would allow us to type coroutines.

@@

Even tough it was quite straightforward to type the MD5 library,
we found a little difference between its documentation and its behavior.
The documentation suggested that the typing of the function \texttt{md5.update}
should be \texttt{(md5\string_context, string) -> (md5\string_context)},
while a comment in the source code suggested that it should be
\texttt{(md5\string_context, string, string*) -> (md5\string_context)}.
However, while testing it and reading its source code, we noticed that
its actual type is \texttt{(md5\string_context, string*) -> (md5\string_context)},
that is, we can pass zero or more strings to \texttt{md5.update}.

This case study shows that it might be not so obvious to identify which
functions discard extra arguments when we are typing an external library.
In this case study, reading its source code and test script was essential
to confirm the typing of \texttt{md5.update}.

>>>>

There are three case studies that we randomly selected from the LuaRocks
repository, they are: HTTP Digest \citep{luahttpdigest},
Typical \citep{luatypical}, and Mod 11 \citep{luamod11}.
The first one provides client side HTTP digest authentication for Lua.
The second one is an extension to the primitive function \texttt{type}.
The third one is a generator and checker of modulo 11 numbers.
We randomly selected these libraries to evaluate Typed Lua for
rewriting existing Lua modules.

@@

\begin{table}[!ht]
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|c|c|}
\hline
\textbf{Case study} & \textbf{easy} & \textbf{poly} & \textbf{inter} & \textbf{hard} & \textbf{Total} & \textbf{\%} \\
\hline
Lua Standard Library & 80 & 6 & 8 & 35 & 129 & 62\% \\
\hline
MD5 & 13 & 0 & 0 & 0 & 13 & 100\% \\
\hline
LuaSocket & 109 & 1 & 2 & 11 & 123 & 89\% \\
\hline
HTTP Digest & 0 & 0 & 1 & 0 & 1 & 0\% \\
\hline
Typical & 1 & 0 & 0 & 0 & 1 & 100\% \\
\hline
Modulo 11 & 7 & 0 & 0 & 2 & 9 & 78\% \\
\hline
Typed Lua Compiler & 144 & 0 & 1 & 9 & 154 & 94\% \\
\hline
\end{tabular}
\end{center}
\caption{Evaluation results for each case study}
\label{tab:evalbycase}
\end{table}

@@

In the following sections we discuss each case study in more detail.
For each case study, we split the evaluation results according
to the modules that each one of them include.
This separation by case study allows us to better discuss the
contributions and limitations of our type system.

@@

All of the modules in the standard library are implemented in C, so
we used Typed Lua to type just the interface of each module.
The \texttt{debug} module is the only one that we did not include in our
evaluation results, because it provides several functions that violate
basic assumptions about Lua code \citep{luamanual}.
For instance, we can use its function \texttt{setlocal} to change the value
of a local variable that is not visible in the current scope.
Table \ref{tab:evallsl} summarizes the evaluation results for the Lua Standard Library
(version 5.3).

\begin{table}[!ht]
\begin{center}
\begin{tabular}{|c|c|c|c|c|c|c|}
\hline
\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{inter} & \textbf{hard} & \textbf{Total} & \textbf{\%} \\
\hline
base & 9 & 1 & 2 & 14 & 26 & 35\% \\
\hline
coroutine & 1 & 0 & 0 & 6 & 7 & 14\% \\
\hline
package & 5 & 0 & 0 & 3 & 8 & 62\% \\
\hline
string & 12 & 0 & 0 & 4 & 16 & 75\% \\
\hline
utf8 & 6 & 0 & 0 & 0 & 6 & 100\% \\
\hline
table & 1 & 5 & 0 & 1 & 7 & 14\% \\
\hline
math & 22 & 0 & 4 & 1 & 27 & 81\% \\
\hline
io & 15 & 0 & 0 & 6 & 21 & 71\% \\
\hline
os & 9 & 0 & 2 & 0 & 11 & 82\% \\
\hline
\end{tabular}
\end{center}
\caption{Evaluation results for Lua Standard Library}
\label{tab:evallsl}
\end{table}

We could give precise static types to several members of
the Lua Standard Library, though some of them are still
hard to type with the current type system.
Now, we discuss why some modules are difficult to type,
and what kind of limitations they represent on our type system.

The \texttt{base} module was quite hard to type because it
includes several functions that have a dynamic behavior.
For instance, the function \texttt{next} traverses all the
fields of a table, but the order that the indices are
enumerated is not specified.
For this reason, it is also difficult to type the function
\texttt{pairs}, as it uses the function \texttt{next} to
traverse a table.
The functions \texttt{getmetatable}, \texttt{rawget}, and
\texttt{rawset} are other examples of functions that their
behavior depend on the C implementation.

We could not give a precise static type to \texttt{ipairs}
because our type system does not have parametric polymorphism.
This function returns an iterator function that traverses a
list of elements.

The \texttt{base} module also includes two special cases:
the functions \texttt{tonumber} and \texttt{collectgarbage}.

The function \texttt{tonumber} is a special case because the
type of the first parameter depends on the type of the
second parameter.
For instance, we can call \texttt{tonumber(1)}, but we cannot
call \texttt{tonumber(1,2)}.
More precisely, the first argument of \texttt{tonumber} can be
a value of any type if it is the only argument, but it must
be a string if there is a second argument,
which must be an integer.

The function \texttt{collectgarbage} is a special case because
its return type depends on the value of the first parameter,
which should be a string.
For instance, calling \texttt{collectgarbage("collect")} returns an integer number,
calling \texttt{collectgarbage("count")} returns a floating point number,
and calling \texttt{collectgarbage("isrunning")} returns a boolean.

We could not type most of the members of the \texttt{coroutine} module
because our type system cannot describe the computational effects of a program.
Lua has one-shot delimited continuations \citep{james2011yield}
in the form of \emph{coroutines} \citep{moura2009rc}, and
effect systems \citep{nielson1999type} are an approach that we
could use to describe control transfers with continuations.
However, for now, coroutines are out of the scope of our type
system, and we use an empty \emph{userdata} declaration
to represent the type \emph{thread}.
Still, we could type the function \texttt{coroutine.isyieldable},
as it has no input parameters, and simply returns a boolean that
indicates whether the running coroutine is yieldable.

@@

Lua programmers often use the functions of the \texttt{string} module
in an object-oriented style, and we treat this feature as a
special case in the implementation.
For this reason, we did not include the typing of the \texttt{string}
methods in our evaluation results.

The \texttt{string} module is not that hard to type, but it includes
some functions that require dependent types.
For instance, the type of the arguments that we pass to
\texttt{string.format} depend on the format string we are using.
It is fine to call \texttt{string.format("\%d", 1)}, but
\texttt{string.format("\%d", true)} generates a run-time error.

The \texttt{utf8} module was quite straightforward to type,
as its members are only operations over strings.

@@

However, the lack of parametric polymorphism did not prevent us from
giving a precise type to \texttt{table.concat}, as it operates over lists
where all elements are strings or numbers.

Even if our type system had parametric polymorphism, it would
be difficult to type \texttt{table.insert}, as its type depends on
the calling arity.
More precisely, calling \texttt{table.insert(l, v)} inserts the
element \texttt{v} at the end of the list \texttt{l},
while \texttt{table.insert(l, p, v)} inserts the element \texttt{v}
at the position \texttt{p} of the list \texttt{l},
but generates a run-time error when \texttt{p} is out of bounds.
This function also does not follow the semantics of Lua on
discarding extra arguments, and generates a run-time error whenever
we pass more than three arguments, even if the first three arguments
match its signature.

Even though the \texttt{math} module looks straightforward to type,
it includes several functions that the return type depend on the input type.
For instance, \texttt{math.abs(1)} returns the integer \texttt{1},
while \texttt{math.abs(1.1)} returns the floating point \texttt{1.1}.

The \texttt{math} module also includes a special case that is hard to type:
the function \texttt{math.random}.
This function is difficult to type because its type depends on the calling arity.
Calling \texttt{math.random()} returns a random floating point number
between \texttt{0} and \texttt{1}.
Calling \texttt{math.random(10)} is equivalent to \texttt{math.random(1,10)},
and returns an integer between this interval.
This function also does not follow the semantics of Lua on discarding extra arguments,
and generates a run-time error whenever we pass more than two arguments.
There is also a problem on its documentation, as it suggests that
the two integers are optional parameters, but, for \texttt{math.random},
optional parameters behave in a different way.
Usually, Lua functions replace optional parameters with a default value
when they are \texttt{nil}, but \texttt{math.random} generates a run-time
error instead.

@@

The function \texttt{io.close} is difficult to type
because its return type depends on whether the file handle that is
being closed was created with \texttt{popen}.
Calling \texttt{io.close(f)} returns the tuple \texttt{(boolean?, string, number)}
when \texttt{f} was created with \texttt{popen},
and it returns the tuple \texttt{(boolean)?} otherwise.
The method \texttt{file:close} is difficult to type for the same reason.

The functions \texttt{io.read} and \texttt{io.lines} are difficult
to type because their return types depend on their input types.
For instance, \texttt{io.read("l")} returns a string or \texttt{nil},
\texttt{io.read("n")} returns a number or \texttt{nil}, and calling
\texttt{io.read("l", "n")} returns a string or \texttt{nil} and a number or \texttt{nil}.
The function \texttt{io.lines}, and the methods \texttt{file:read} and
\texttt{file:lines} are difficult to type for the same reason.

The function \texttt{os.execute} is the only member of the \texttt{os}
module that we could not give a precise type.
This function is difficult to type because its return type depend on
the input type.
More precisely, \texttt{os.execute()} returns a boolean, while
\texttt{os.execute(c)} returns \texttt{(boolean?, string, number)},
where \texttt{c} is a string that represents the command to be executed.

The evaluation results show that our type system should include
intersection types, parametric polymorphism, and effect types,
as these features would help us increase the static typing of the
Lua Standard Library.
Intersection types would allow us to define overloaded function types.
Parametric polymorphism would allow us to define generic function and table types.
Effect types would allow us to type coroutines.

@@

Even tough it was quite straightforward to type the MD5 library,
we found a little difference between its documentation and its behavior.
The documentation suggests that the type of the function \texttt{md5.update}
is \texttt{(md5\string_context, string) -> (md5\string_context)},
though there is a call to this function in the test script that passes
an extra string argument.
Reading the source code, we found that its actual type is
\texttt{(md5\string_context, string*) -> (md5\string_context)},
that is, we can pass zero or more strings to \texttt{md5.update}.

This case study shows that type annotations help programmers maintain the
documentation updated, as the type checker always validate them.
